---
title: "data609hw3"
author: "jim lung"
date: "September 15, 2018"
output: html_document
---

__113__

2) The following table gives the elongation $e$ in inches per inch (in./in.) for a given stress S on a streel wire measured in pounds per square inch ($lb/in^2$). Test the model $e=c1S$ by potting the data. Estimate $c1$ graphically.


```{r}
library(ggplot2)
```


```{r}
S <- c(5, seq(10, 100, 10))
e <- c(0, 19, 57, 94, 134, 173, 216, 256, 297, 343, 390)

qplot(x = S, y = e)
```

Looks like a straight line. At about y = 100 it should be at about x = 30, so the slope linear model has a slope of about 3. 

__121__

2a) For each of the following data sets, formulate the mathematical model that minimizes the largest deviation between the data and the line y = ax+b. If a computer is available, solve for the estimates of a and b.

```{r, tidy=TRUE}
x <- c(1.0,2.3,3.7,4.2,6.1,7.0)
y <- c(3.6,3.0,3.2,5.1,5.3,6.8)

df <- data.frame(x,y)

lm <- lm(y ~ x, df)
lm$coefficients

f3 <- ggplot(data=df) + geom_point(aes(x=x, y=y)) + geom_abline(intercept= 2.214853, slope=0.5642337)
f3
```


__127__

10) Fit the data with the models given using least squares. Data for planets. Fit the model $y=ax^(3/2)$

```{r}
body <- c("Mercury", "Venus", "Earth", "Mars", 
          "Jupiter", "Saturn", "Uranus", "Neptune")
period <- c(7.6 * 10^6, 1.94 * 10^7, 3.16 * 10^7, 
            5.94 * 10^7, 3.74 * 10^8, 9.35 * 10^8, 
            2.64 * 10^9, 5.22 * 10^9)
distance <- c(5.79 * 10^10, 1.08 * 10^11, 1.5 * 10^11, 
              2.28 * 10^11, 7.79 * 10^11, 1.43 * 10^12, 
              2.87 * 10^12, 4.5 * 10^12)
df <- data.frame(body, period, distance)
df
```

Least-squares equation for power curve where $n=3/2$, x=period and y=distance.

```{r}
n <- 3/2

model <- function (x, y, n)
{
  numerator <- sum(y * x^n)
  denominator <- sum(x^(2*n))
  print(numerator)
  print(denominator)
  result <- numerator / denominator
  return (result)
}

a <- model(df$period, df$distance, n)
a

planet <- function(a, x, n)
{
  yestimate <- a * (x^(n))
  return(yestimate)
}

x_vals <- seq(min(df$period), max(df$period), by=100000)
y_mest <- planet(a, x_vals, n)
dfplanet<- data.frame(x_vals, y_mest)

```

Original data points and the model plotted.

```{r}
f4 <- ggplot(df) + geom_point(aes(x=period, y=distance)) + geom_line(data=dfplanet, aes(x=x_vals, y=y_mest)) 
f4
```

Squared deviations may be minimized, but the model does not do a good job representing the data.